---
title: "a) Project notes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{a) Project notes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = F
)
```

```{r setup, eval = T}
library(compbio4all)
```


## Review biology through charts

chargaff1951

Higgs 2009 amino acids


## TODO

go through all my repos to look for addiationl code

blast-stats project
https://github.com/brouwern/blaststats

https://github.com/brouwern/carnivorephylo

https://github.com/brouwern/CaseStudiesEcoStats


# Packages that could be worth learning

## Formatting output stats: xplain
http://www.zuckarelli.de/xplain/index.html
http://www.zuckarelli.de/xplain/xplain_cheatsheet.pdf



## Processing images: magick
https://cran.r-project.org/web/packages/magick/vignettes/intro.html


This has a lot of dependencies and can take a long time to set up.
```{r eval = F}
install.packages("magick",dependencies = T)
```

Here's an example of how to resize some png files used in this book
```{r}
#get names of files
files <- list.files(path = "./images/Medley1998csv",pattern = "Medley1998",full.names = T)

#make sure i just have png files, not the ppt
files <- files[grep("png",files)]

#look at file
print(image_read(files[1]))

newfiles <- gsub("Medley1998csv-","Medley1998csv-800x",files)

for(i in 1:length(files)){
  image_write(image_resize(image_read(files[i]), "800x")
, path = newfiles[i]
, format = "png")
}

```



## Data sets 
ChickWeight   Weight versus age of chicks on different diets
chickwts 	Chicken Weights by Feed Type
CO2 	Carbon Dioxide Uptake in Grass Plants
co2 	Mauna Loa Atmospheric CO2 Concentration
InsectSprays   Effectiveness of Insect Sprays
iris 	Edgar Anderson's Iris Data
iris3 	Edgar Anderson's Iris Data
islands 	Areas of the World's Major Landmasses
LakeHuron   Level of Lake Huron 1875-1972
Loblolly   Growth of Loblolly pine trees
lynx 	Annual Canadian Lynx trappings 1821-1934
npk   Classical N, P, K Factorial Experiment
Orange   Growth of Orange Trees
OrchardSprays 	Potency of Orchard Sprays
PlantGrowth   Results from an Experiment on Plant Growth
precip 	Annual Precipitation in US Cities
rivers   Lengths of Major North American Rivers
rock 	Measurements on Petroleum Rock Samples
trees   Girth, Height and Volume for Black Cherry Trees
women   Average Heights and Weights for American Women


http://www.zoology.ubc.ca/~whitlock/ABD/teaching/datasets.html
