---
title: "b) Motivating examples: Species distribution models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{b) Motivating examples: Species distribution models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = T}
library(compbio4all)
```


## Mangroves


R code:
https://figshare.com/articles/dataset/Supplement_1_R_script_for_species_and_community_distribution_models_described_in_the_main_text_/3563811?file=5637201

note: they source 2 files while are missing

All data files:
https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF022

## Origina of new anopholes in Africa

"Phylogenetic analyses of Ethiopian (Kebri Dehar) and Djibouti
specimens reveal that these two populations are from different
clades that originate from different sources in Pakistan
(19)."

T. E. Carter et al., First detection of Anopheles stephensi Liston, 1901 (Diptera: Culicidae)
in Ethiopia using molecular and morphological approaches. Acta Trop. 188,
180–186 (2018).

## Anopholes spread from India to Horn of Africa

A new malaria vector in Africa: Predicting the expansion range of Anopheles stephensi and identifying the urban populations at risk

https://datadryad.org/stash/dataset/doi:10.5061/dryad.3xsj3txcx

Predicting coastal algal blooms with environmental factors by machine learning methods

## Red tide

A Review of Recent Machine Learning Advances for Forecasting Harmful Algal Blooms and Shellfish Contamination

Predicting coastal algal blooms with environmental factors by machine learning methods

## Murder hornets

Geographic potential of the world’s largest hornet, Vespa mandarinia Smith (Hymenoptera: Vespidae), worldwide and particularly in North America

DATA and CODE

https://kuscholarworks.ku.edu/handle/1808/30602

https://github.com/townpeterson/vespa


## Other studies

Geographic abundance patterns explained by niche centrality hypothesis in two Chagas disease vectors in Latin America


## notes

### other methods besides maxent

Classification and regression with random forests as a standard method for presence-only data SDMs: A future conservation example using China tree species

The use of classification and regression algorithms using the random forests method with presence-only data to model species’ distribution

Brazilian Forest Dataset: A new dataset to model local biodiversity

Modelling species presence-only data with random forests

Use of Machine Learning (ML) for Predicting and Analyzing Ecological and ‘Presence Only’ Data: An Overview of Applications and a Good Outlook

Machine learning of large‐scale spatial distributions of wild turkeys with high‐dimensional environmental data
