---
title: "vegan PCA: Principal Components Analysis with vegans rda function"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vegan PCA: Principal Components Analysis with vegans rda function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = T}
library(compbio4all)
```



PCA (Principal Components Analysis) is easy in R, but the standard biplot() function is a little clunky.  The vegan package can do PCA using the rda() function (normally for redundancy analysis) and has some nice plotting functions.  (Note that ggplot is also developing biplot tools).


# Fisher's Irises

* 3 species
* 4 response variables

```{r}
data("iris")
summary(iris)
```

# Visualize variables in 2D

```{r}

pairs(iris,
      lower.panel = NULL, 
      col = as.numeric(iris$Species))
```

Note: can also be done in ggplot using GGalley::ggpairs; however, not updated for most recent version of R.


<br><br>

# Do PCA with base R function

## Run PCA

* NOTE: there is also a similar function to princomp(), prcomp()

```{r}
#"[,-5]" drops that last columns
##the species code
iris.pca <- princomp(iris[,-5])
```

## Plot Biplot in base graphics

```{r}
biplot(iris.pca)
```

There are is also ggplot biplot code somewhere outthere on the internet.

# Do PCA with vegan::rda

* rda = redundancy analysis
* normally RDA is used for "constrained ordination" (ordination w/covariates or predictor)
* without predictors, RDA is the same as PCA
* in vegan, givig the function rda() dataframe without predictors runs a PCA very simiar to princomp


## run a vegan PCA with rda()

```{r}
library(vegan)
my.rda <- rda(iris[,-5])


#"[,-5]" drops that last columns
##the species code
```


<br><br>


## Plot the RDA

Note to get help when doing and RDA biplot, call ?biplot.rda; however, use biplot() for the actual plotting.

### Default biplot of RDA output

```{r}
biplot(my.rda)
```

Note that vegan's scaling is different than a normal biplot.

<br><br>

### biplot.rda options

* vegan is package for ecological analysis and so its functions are oriented towards particular kinds of analyses
* the "display" arguement is not in the normal biplot comman used with princomp
    + "display = 'sites' " corresonds to the rows of data; one datapoint per row
    + "display = 'species'" corresponds to the columns of data; one line/arrow per column
* "type" is an argument also not in biplots originating from princomp
    + "type = 'text' " plots row numbers and column names as labels
    + "type = points' " plots points

```{r}
biplot(my.rda,
       display = c("sites", 
                   "species"),
       type = c("text",
                "points"))
```



# vegan's ordination plotting functions

* vegan has lots of helpful functions for plotting ordinations.
    + ordihull: convex polygons around groups of points
    + ordiellipse: ellipses around centroids (SD or SE)

## Add "hulls" around each species
```{r}
#make basic plot
biplot(my.rda,
       display = c("sites", 
                   "species"),
       type = c("text",
                "points"))

#Add "hulls"
ordihull(my.rda,
         group = iris$Species)
```



## Make plot pretty

```{r}
#make basic plot
biplot(my.rda,
       display = c("sites", 
                   "species"),
       type = c("text",
                "points"))

#get names of species
## levels() extracts fator levels
## from the column
spp.names <- levels(iris$Species)

#Add hulls
ordihull(my.rda,
         group = iris$Species,
         col = c(1,2,3))


#Add legend
legend("topright",
       col = c(1,2,3), 
       lty = 1,
       legend = spp.names)
```





