---
title: "Anscomb's quartet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Anscomb's quartet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = T}
library(compbio4all)
```







# The Data

## x variables
```{r}
#x1 through x3 are the same
x1 <- c(10,8,13,9,11,14,6,4,12,7,5)
x3 <- x2 <- x1

#x4
x4 <- c(8,8,8,8,8,8,8,19,8,8,8)


```


## y varibles are all different
```{r}
y1 <- c(8.04,6.95,7.58,8.81,8.33,9.96,7.24,4.26,10.84,4.82,5.68)

y2 <- c(9.14,8.14,8.74,8.77,9.26,8.1,6.13,3.1,9.13,7.26,4.74)

y3 <- c(7.46,6.77,12.74,7.11,7.81,8.84,6.08,5.39,8.15,6.42,5.73)

```


<br><br>

# Plots

## Plot y1 vs x1

```{r}
plot(y1 ~ x1, 
     main = "y1 ~ x1")
```


## Adjust y and x limits

* arguments "xlim ="" and "ylim = "

```{r}
plot(y1 ~ x1, 
     main = "y1 ~ x1",
     xlim = c(0,15),
     ylim = c(0,14))
```




## Calcualte correlation of y1 and x1
```{r}
cor(x1,y1)
```

## Linear regression y1 ~ x1

* Save output to object "y1x1"
* Not "data = " is required b/c the data are living in seperate vectors stored directly into R's memory

<br>

### Do linear regression
```{r}
y1x1 <- lm(y1 ~ x1)
```

<br>

### Look at output

```{r}
summary(y1x1)
```


### Plot data with regression line

* abline() add regression line
* col = changes color
* lwd = changes line width
* pch = 16 changes plotting symbol
* cex = 2 increass point size

```{r}
#scatter plot
plot(y1 ~ x1, 
     main = "y1 ~ x1",
     xlim = c(0,15),  #axis limits
     ylim = c(0,14),
     pch = 16,        #pnt shape
     cex = 2)         #pnt size

#regression line
abline(y1x1, 
       col = 2, #line color
       lwd = 3) #line width
```


## Add R^2^ values

* get R^2^ by saving output of summary command
    + y1x1r2 <- summary(y1x1)$r.squared
* use mtext() to add annotation of R^2^2

```{r}
#scatter plot
plot(y1 ~ x1, 
     main = "y1 ~ x1",
     xlim = c(0,15),  #axis limits
     ylim = c(0,14),
     pch = 16,        #pnt shape
     cex = 2)         #pnt size

#get r.squared
y1x1r2 <- summary(y1x1)$r.squared
y1x1r2 <- round(y1x1r2,3)

#annotation
mtext(text = y1x1r2)

#regression line
abline(y1x1, 
       col = 2, #line color
       lwd = 3) #line width
```

<br><br><br>

# Make two side by side plots

* 1^st^, add another set of data, ie y2 and x2, etc
* Do regression on new data
    + Look at coefficients and R^2^
* Plot side by side with 1st data
    + use par(mfrow = c(1,2)) to set up a 1 x 2 panel
* NOTE: both plots much be in same Rmarkdown "chunk"


```{r}
#SET up 1 x 2 panel
par(mfrow = c(1,2))

#PLOT 1
##scatter plot 1
plot(y1 ~ x1, 
     main = "y1 ~ x1",
     xlim = c(0,15),  #axis limits
     ylim = c(0,14),
     pch = 16,        #pnt shape
     cex = 2)         #pnt size

#get r.squared
y1x1r2 <- summary(y1x1)$r.squared
y1x1r2 <- round(y1x1r2,3)

#annotation
mtext(text = y1x1r2)

#regression line 1
abline(y1x1, 
       col = 2, #line color
       lwd = 3) #line width


#PLOT2

## regression y2 ~ x2
y2x2 <- lm(y2 ~ x2)

##scatter plot 2
### note: I set col = 0 to hide output
### deletc this or change it to 1
plot(y2 ~ x2, 
     main = "y2 ~ x2",
     xlim = c(0,15),  #axis limits
     ylim = c(0,14),
     col = 0,
     pch = 16,        #pnt shape
     cex = 2)         #pnt size

#get r.squared for 2nd model
y2x2r2 <- summary(y2x2)$r.squared
y2x2r2 <- round(y2x2r2,3)

## 2nd R^26 annotation
### note: I set col = 0 to hide output
### deletc this or change it to 1
mtext(text = y2x2r2, 
      col = 0)

## 2nd regression line
### note: I set col = 0 to hide output
### deletc this or change it to 1
abline(y2x2, 
       col = 0, #line color
       lwd = 3) #line width


```

