---
title: "A primer of population ecology: Part 2 - How do populations change?"
author: "Nathan Brouwer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A primer of population ecology: Part 2 - How do populations change?}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval  = F,
  comment = "#>"
)
```


## Key vocab

Put key vocab here


## Populations can change their size, density, or spatial extent

We can consider changes in populations from multiple angles  For example, Kirkland's Warbler is currently

1. Increasing in overall number of individuals.
1. Increasing in number of occupied habitats. 
1. Increasing its species range.

Also, since the warbler prefers a certain density of Jack Pine, its density also changes.  Jack Pine stands frequently burn in **forest fires**, and are also logged for timber.  As the density of trees changes, the density of warblers changes.


## Many studies and models of population growth focus on changes in population size

Discussions of population dynamics often center on changes in population size.  Changes in population size are often plotted in a **time series graph** with time on the x-axis (usually in years) and population size (N) on the y-axis.  **Patterns of population change** in terms of population size include 

1. **Growth**: growing larger than the current size (Snail kites: Figure 1)
1. **Decline**: decreasing in abundance (Elk: Figure 2)
1. **Stability**: staying approximately the same size over time (Wolves: Figure 3)
1. **Recovery**: Stability or growth following a period of decline. (Impala: Figure 4).
1  **Variation**: repeatedly varying between growth and decline. 
1. **Extirpation (local extinction):** decline of one population of a species to 0.  (Kirkland's warbler: Figure 7)
1. **Extinction**: decline of all members of a species to 0. (NO EXAMPLE)
1. **Cycles**: repeated patterns of growth followed by decline (Lynx: Figure 8)

```{r, ech = F}
#Growth
## SYKES
#RECENT POPULATION TREND OF THE SNAIL KITE IN
# FLORIDA AND ITS RELATIONSHIP TO WATER LEVELS
yr.kite <- c(1969:1980)
n.kite <- c(98,120,72,65,95,81,110,142,152,267,431,651)
plot(n.kite ~ yr.kite, type = "b",
     xlab = "Year",
     ylab = "Population size",
     main = "Growth: Snail kites, Florida")

#Decline
### Elk
# Peek et al. 1967 Population Changes within the Gallatin Elk Herd, 1932-65
yr.max <- c(1932,1936,1944,1955,1964)
n.max <- c(2475,2218,2178,2177,2145)
plot(n.max ~ yr.max, type = "b",
     xlab = "Year",
     ylab = "Elk",
     main = "Decline: Gallatin elk herd, Montana")
# yr.elk <- c(1932,1936,1944,
#             1948:1964)
# n.elk  <- c(2475,2218,2178,
#            1372,1709,1570,1603,1316,2106,1842,2177,1235,1269,1221,1379,1688,1430,1470,1656,2145)
# 
# plot(n.elk ~ yr.elk, type = "b")


## Wolves
# Peterson et al 
# POPULATION LIMITATION AND THE WOLVES OF ISLE ROYALE
#this just shows stability after a crash
yr.wolves <- c(1987:1995)
n.wolves <- c(16,12,12,15,12,12,13,17,17)
plot(n.wolves ~ yr.wolves,
     type = "b",
     ylim = c(0,20),
     xlab = "Year",
     ylab = "Wolves",
     main = "Stability: Isle Royale Wolves")

# Impala
# 
yr.imp <- c(1959,
            1965,1967,1968,1975,1981,1982,1983,
            1984, 1985, 1991)
n.imp  <- c(1500,#estime
            285,300,451,985,1332,1436,1576,
            212,300,1500)
plot(n.imp ~ yr.imp, type = "b",
     xlab = "Year",
     ylab = "Population size",
     main = "Recovery: Impala Lake Manyara
 National Park & Anthrax Outbreaks",
 ylim = c(190,2200))
abline(v = 1961, lty = 2, lwd = 2)
abline(v = 1977, lty = 2, lwd = 2)
abline(v = 1984, lty = 2, lwd = 2)
legend(x = "topleft",legend = "Anthrax outbreaks",lty = 2, lwd = 2)

## Extirpation
# Probst 1986 A Review of Factors Limiting the Kirtland's Warbler on its Breeding Grounds
yr.KIWA          <- c(1972:1982)
N.KIWA.artillery <- c(50,48,25,24, 8, 5,0,0,0,0,0)
N.KIWA.ogemaw    <- c(48,49,26,23,23,10,5,0,0,0,0)
plot(N.KIWA.artillery ~ yr.KIWA, type = "b", xlab = "Year", ylab = "Singing KIWA males",
     ylim = c(0,50),
     main = "Extirpation: Kirkland's Warbler in Two Michigan Forests")
points(N.KIWA.ogemaw ~ yr.KIWA, type = "b", pch = 16, lty = 3, col = 2)
abline(h = 0)

# Extinction Northern white rhino
#https://en.wikipedia.org/wiki/Northern_white_rhinoceros#cite_note-Times-38
yr.nwr <- c(1909, 1960, 1975,1984,1995,1996,1998,2000,2003,2006,
2007, 2008, 2009, 2011,2014, 2015, 2017,2018)
n.nwr <- c(2500,2000,500,15,35,29,28,33,30,4,0,0,0,0,0,0,0,0)
n.nwr.cap <- c(0,7,15,16,11,10,10,11,11,10,8,8,8,7,5,3,3,2)
plot(n.nwr ~ yr.nwr, type = "b",
     xlab = "Year",
     ylab = "Rhinos",
     main = "Extinction: Northern white rhinos")
points(n.nwr.cap ~ yr.nwr, type = "l")

### Cycles
data("lynx")
plot(lynx,main = "Cycles: Lynx population cycles",
     xlab = "Year",ylab = "Number of Lynx trapped")
```


The **processes** and **mechanisms** that drive population change can include availability of resources, weather, inbreeding, predators or parasites.  In this reading, we will focus mostly on **patterns** of change and how they are described mathematically, and won't focus much on the processes.

For example, the Kirkland's Warblers (*Setophaga kirtlandii*; KIWA for short) is a species of **migratory** songbird **endemic** to North America.  KIWA populations are monitored by determining the number of males defending territories in their summer breeding habitat in the Great Lakes, primarily Michigan.  There were about 500 males with territories in the 1950s (Figure 1).  The following changes occurred over the next 50 years:

1. **Decline** over the course of the 1960 to ~200 territories.
1. A period of **stability** at ~200 territories from 1975 to 1990.
1. Steady **growth** to >2500 from 1990 through 2020.



```{r, echo = F}

# 1023 within the forest, estimating it to be "greater than 40%" of the #total population in the state. 
#2,558 warblers #for 2019
#www.fs.usda.gov/inside-fs/delivering-mission/sustain/forest-service-census-reveals-healthy-kirtlands-warbler
# 40/100 = 1023/x
# 0.40*x = 1023
#1023/0.40

N <- c(432,502,201,200,216,167,179,200,219,200,211,243,232,207,215,215,216,210,167,207,212,265,347, 397,485,633,765,692,728,804,904,891,1085,1050,1202,1341,1420,1485,1707,1803,1826,1773,1828,2090,2020, NA,2365,
       1023/0.40)

year <- c(1951,1961,1971,1972,1973,1974,1975,1976,1977,1978,1979,1980,1981,1982,1983,1984,
          1985,1986,1987,1988,1989,1990,1991,1992,1993,1994,1995,1996,1997,1998,1999,2000,
          2001,2002,2003,2004,2005,2006, 2007,2008,2009,2010,2011,2012,2013,2014,2015,
          2019)

plot(N ~ year, type = "b", main = "Kirkland warbler abundance, 1951-2015",
     xlab = "Year", ylab = "Number of singing males",ylim = c(0,3000))




```

Changes in population size over time and the processes that cause these to occur are called **population dynamics**.  How populations change in abundance over time is a major concern of **population ecology**, **wildlife ecology**, and **conservation biology**, and shares many similarities with questions asked in **evolutionary biology**.  

Researchers who study population dynamics often use **mathematical models** to describe and predict population dynamics and understand what factors are driving those changes. For example, if there are 2500 Kirkland's Warblers in Michigan this year, can be predict how many will be around next year, or 10 years from now?  Due to its small population size KIWA was listed as an Endangered Species in 1967.  In 2019 it was de-listed and now is considered "Near-threatened."  Ecologists are very interested in using models to predict how large the KIWA population will be in the future, and what factors cause it to increase and decrease.

