---
title: "R quirks: putting things in vectors"
author: "Nathan Brouwer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{R quirks: putting things in vectors}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```



# How R works with strings of numbers


# How R works with strings of numbers

R has a somewhat strange behavior.  Look at this example where we are trying to find the mean of 3 numbers, 0.86, 1.02, and 1.02.

It might be tempting to calculate the mean of these three numbers like this.
```{r}
mean(0.86,1.02,1.02)
```

Notice, however, that the value that gets returned, 0.86, is exactly the same as the 1st value in the string of numbers.  Hmmmmm.... This is definitely not the mean.

Now try this: put the numbers into a "vector" using the functionc c()

```{r}
y <- c(0.86,1.02,1.02)

```

We can check that the numbers are there
```{r}
y
```


And we can get the mean using the mean command
```{r}
mean(y)
```


This seems like a more reasonable answer, but since R seems to be acting weird we can check our answer another way
```{r}
sum(y)/length(y)
```

Ok, seems ok.  So whats wrong with the origin way, of just running "mean(0.86,1.02,1.02)" ?  

In general, R functions carry out operations on **vectors**, or sets/strings of nubmers that have been packaged together using the **c()** command or contained in the column of a dataframe or matrix.  The function mean() is expecting to get a vector, not a list of raw data.  Technically, what its doing is interpreting the value 0.86 as a vector, taking its mean (which is 0.86) and ingoring the rest of the numbers.  Unfortuantely, it isn't giving you a warning message.


# Fixing the problem

We can get our oringal bit of code to work by just putting the data 0.86,1.02,1.02 into a vector using c(), like this
```{r}
mean(c(0.86,1.02,1.02))
```

This is just a more direct way of doing the following, just using a single line of code instead of two.
```{r}
y <- c(0.86,1.02,1.02)
mean(y)
```





