---
title: "Basic statistics equations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic statistics equations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = T}
library(compbio4all)
```


## Variance

\[\sigma^2 = {\frac{\sum\limits_{i=1}^{n} \left(x_{i} - \bar{x}\right)^{2}} {n-1}}\]

## Standard deviation

\[\sigma = \sqrt{\frac{\sum\limits_{i=1}^{n} \left(x_{i} - \bar{x}\right)^{2}} {n-1}}\]



## Standard error

### Equation
\[SE = \frac{SD}{\sqrt{N}}\]

### R-ish notation
\[SE = \frac{SD(x)}{sqrt(length(x))}\]


## Pooled variance for two-sample t-test
In words in terms of variance:
\[var_p = {\frac{df_1*var_1 + df_2*var_2}{df_1 + df_2}}\]

In words in terms of standard deviation (SD)
\[var_p = {\frac{df_1*SD^2_1 + df_2*SD^2_2}{df_1 + df_2}}\]

In symbols, in terms of the variance (sigma^2)
\[\sigma^2_p = {\frac{df_1*\sigma^2_1 + df_2*\sigma^2_1}{df_1 + df_2}}\]



## Standard error of the difference
\[SE_diff = \sqrt{{s_p^2}(\frac{1}{n_1} + \frac{1}{n_2})} \]


## LS regression slope

* NOTE: squaring and multiplicatino occur before summing
    + therefore for denomiator square each observation 1st, then sum them all up

### Words
\[slope = \frac{\sum (X deviations)*(Y deviations)}{\sum (X  deviations^2)} \]


### Math
\[b = \frac{\sum (X_i-\bar{X})*(Y_i-\bar{Y})}{\sum (X_i-\bar{X}^2)} \]



# Regression intercept

* Your mean Y value is related to your mean x value like this

## Word equation
\[\bar{Y} = Intercept + Slope*\bar{X} \]

<br>

## Math equation
\[\bar{Y} = a + b*\bar{X} \]

<br>

## Do some algebra to solve for the intercept a
\[a = b*\bar{X} - \bar{Y}\]






# F statistics (F ratio)







## Chi^2 test

(I just copied and pasted this and have not checked it)


\[{\chi ^{2}=\sum _{i=1}^{n}{\frac {(O_{i}-E_{i})^{2}}{E_{i}}}=N\sum _{i=1}^{n}{\frac {\left(O_{i}/N-p_{i}\right)^{2}}{p_{i}}}}\]




## Power


\[Power \propto \frac{ES*\alpha*\sqrt{n}}{\sigma^2}\]


