---
title: "waRmup: entering data & making R objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{waRmup: entering data & making R objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = T}
library(compbio4all)
```







We'll take some simple data from Motulsky, enter it into R by hand, and do a t-test.  This will require making two R "objects" to hold the data for us.

# R Functions

* "<-": the assignment operator
* "c(...)": the concatenate function; holds stuff, like numeri data
* "t.test": does a t-test!

<br><br>

# The Data

Motulsky 2nd Ed, Chapter 30, page 220, Table 30.1.  Maximal relaxaction of muscle strips of old and young rat bladders stimualted w/ high concentrations of nonrepinephrine (Frazier et al 2006).  Response variable is %E.max

```{r, echo = F}
df.rats <- data.frame(old = c(20.8,2.8,50.0,33.3,29.4,38.9, 29.4,52.6,14.3),
           young = c(45.5,55.0, 60.7, 61.5, 61.1, 65.5,42.9,37.5, ""))
```


```{r, echo = F}
library(knitr)
kable(df.rats, caption = "Motulsky Table 30.1 Rat bladders")
```

# Task 1: Load data

Load each column of data into a seperate R object

HINTS: You'll need need to use:

* <-
* c(...)
* don't forget commas!
* don't put spaces or dashes w/in the names of R objects; use periods or undescores
* an.object.name
* an_boject_name
* Scroll down for more hints

```{r}
#TYpe your attempt here:





```


<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

HINT: The basic form of the code is:
* R.object <- c(datum1, )

```{r}
#Type your attempt here:




```



<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

HINT: To make an object with just the first "old" datum do this

```{r}
old.E.max <- c(20.8)
```

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

HINT: To make another object with just the first "young" datum do this

```{r}

young.E.max <- c(45.5)

```

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

## Loaded data

```{r}

#Old data
old.E.max <- c(20.8,2.8,50.0,33.3,29.4,38.9, 29.4,52.6,14.3)

#Young data
young.E.max <- c(45.5,55.0, 60.7, 61.5, 61.1, 65.5,42.9,37.5)

```


<br><br>

# T-test on rat bladders

Perform a t-test to compare the means of these 2 samples.  This requires

* the "old.E.max" object (or whatever you called it)
* the "young.E.max"
* the function t.test()
* a single comma

Scroll down for the answer...

```{r}
#Type your attempt here:




```


<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

```{r}

t.test(old.E.max, young.E.max)

```

