---
title: "z0) Undergraduate editors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{z0) Undergraduate editors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = F}
library(compbio4all)
```



## Winter 2021

Aman Virmani AMV88@pitt.edu
Noah Burget NGB23@pitt.edu
Emily Jenison EMJ45@pitt.edu
Havannah Tung. <LET61@pitt.edu>
Nitish?


## Original students emailed
	"Burget, Noah G" <ngb23@pitt.edu>,
"Whitfield, Tyler S" <Tsw32@pitt.edu>,
"Kalepalli, Nishita" <nik78@pitt.edu>,
ARR130@pitt.edu,
"Sheng, Tao" <tao_sheng@pitt.edu>,
"Shetty, Shriya S" <SSS92@pitt.edu>,
"Cherupally, Vijay Kiran" <VIC38@pitt.edu>,
"Cheng, Claire A" <cac332@pitt.edu>,
"Murray, Anna M" <Amm498@pitt.edu>,
"Virmani, Aman" <amv88@pitt.edu>,
brd98@pitt.edu,
"Varghese, Christopher G" <cgv7@pitt.edu>,
"Morley, Lucas N" <Lnm48@pitt.edu>,
"Prem, Michelle" <mrp105@pitt.edu>,
"Fiore, Jack A" <Jaf192@pitt.edu>,
"Shetty, Saumya S" <sss93@pitt.edu>,
"Chamania, Nitish" <nic64@pitt.edu>,
"Walbridge, Jordan H" <jhw43@pitt.edu>,
"Hutchins, Theresa R" <Trh53@pitt.edu>,
thk44@pitt.edu,
"Reich, Isaac J" <ijr11@pitt.edu>,
"Walsh, Jake" <jaw312@pitt.edu>,
"Spellman, David A" <das320@pitt.edu>,
"Ramaradj, Priyanka" <prr44@pitt.edu>,
"Tung, Havannah" <Let61@pitt.edu>,
"Katz, Tyler J" <TJK68@pitt.edu>,
"Shi, Tingjia" <tis46@pitt.edu>,
"Sharoubeem, Paula" <Pas195@pitt.edu>,
"Khamar, Ronak K" <rkk19@pitt.edu>,
"Wang, Li-Ting" <liw89@pitt.edu>,
"Jenison, Emily Grace" <emj45@pitt.edu>,
"Dowling, Rileigh" <rid32@pitt.edu>



## Original email

A spreadsheet where you can sign up for tasks is here:
https://docs.google.com/spreadsheets/d/1MjUMNidAHfTTHrF0p7MjOUysDC-LGAz3RYgbhJJWGAc/edit?usp=sharing


A book which I want to integrate into "Comp Bio For All!" but haven't done any work on yet is
https://little-book-of-r-for-multivariate-analysis.readthedocs.io/en/latest/

I'd like to integrate the sections below into CB4ALL.  The main task here is turning these webpages into .rmd files with the code in code chunks, section titles formatted, etc.

https://little-book-of-r-for-multivariate-analysis.readthedocs.io/en/latest/src/multivariateanalysis.html#plotting-multivariate-data
https://little-book-of-r-for-multivariate-analysis.readthedocs.io/en/latest/src/multivariateanalysis.html#calculating-summary-statistics-for-multivariate-data
https://little-book-of-r-for-multivariate-analysis.readthedocs.io/en/latest/src/multivariateanalysis.html#calculating-correlations-for-multivariate-data
https://little-book-of-r-for-multivariate-analysis.readthedocs.io/en/latest/src/multivariateanalysis.html#standardising-variables
https://little-book-of-r-for-multivariate-analysis.readthedocs.io/en/latest/src/multivariateanalysis.html#principal-component-analysis 

I've worked on most of this book
https://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/

but there are several chapters I have not yet done anything with.  These chapters need to be pasted into .rmd, code put into code chunks, etc.
https://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter10.html
https://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/revisionexercises_answers.html
https://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter11.html

All graphs in the book are currently done with base R graphics,  For any chapter, including the ones I've worked on you can re-work the graphs using ggpubr or ggplot2.
