---
title: "TODO"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TODO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = F}
library(compbio4all)
```

https://rdrr.io/cran/NameNeedle/man/needles.html

https://open.oregonstate.education/appliedbioinformatics/chapter/chapter-3/

https://farhanma.github.io/pyseq/

https://iubmb.onlinelibrary.wiley.com/doi/full/10.1002/bmb.21027

http://www.cs.bilkent.edu.tr/~calkan/teaching/cs481/pdfslides/07-needlemanwunsch-smithwaterman.pdf

http://rna.informatik.uni-freiburg.de/Teaching/index.jsp?toolName=Needleman-Wunsch

https://wilkelab.org/classes/SDS348/2019_spring/labs/lab13-solution.html

https://www.biostars.org/p/146980/

https://gist.github.com/juliuskittler/ed53696ac1e590b413aac2dddf0457f6


machine-learning using sequenceds
https://academic.oup.com/bioinformatics/article/34/14/2499/4924718
https://academic.oup.com/bib/article/21/3/1047/5475015
https://cran.r-project.org/web/packages/ftrCOOL/ftrCOOL.pdf
https://ilearn.erc.monash.edu/
https://www.pnas.org/content/92/19/8700
https://academic.oup.com/bioinformatics/article/33/1/122/2525676

https://www.nature.com/articles/nrg3920
https://academic.oup.com/bioinformatics/article/31/2/279/2365812?login=true


1. Copy text and code from original website to .rmd files
1. Move any functions AC wrote to .R files
1. Move code into R code chunks
1. Compile list of packages used in the chapter; make sure these are loaded at beginning of chapter
1. Compile list of AC functions used in the chapter
1. Compile list of dad used in the chapter
1. Add AC functions to biodata package
1. Add dat to the biodata package
1. Make sure biodata gets called at beginning of chapter to load any ASC

